Magic-free Code Extraction of Generic Programs and Recursion Schemes in Coq

### 2024-03-18 paper/paper.tex
anamorphisms

\subsection{Coalgebras and Anamorphisms}
\label{sec:coalg}
While folds \emph{destruct} data, their dual, the unfolds, \emph{construct}
data. In particular, given a seed value and an \emph{observation}, the unfold
corecursively  ``runs'' the observations starting from the initial seed.
% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
% \[\begin{tikzcd}
%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
% 	\arrow["a", from=1-1, to=1-2]
% 	\arrow["b", from=1-2, to=1-3]
% \end{tikzcd}\]
The seed function can be modelled by a coalgebra. For example, for a type of
states \coq{X} and an alphabet type \coq{L} we can define a (labelled)
transition system (LTS) on \coq{X} as a function \coq{c : X -> L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.
In our development we use the following notation for coalgebras:
\begin{coqcode}
Notation Coalg F A := (A ~> App F A).
\end{coqcode}
Now an anamorphism takes such a seed function and runs it while collecting all
the observable behaviours. In order to do so we define the type of infinite
streams of labels which is the greatest fixed-point for the functor \coq{A * X}.
We define this in Coq as follows:
\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
where \coq{GFix_in}  and \coq{GFix_out} are the two inverses witnessing the
isomorphism. Similarly to \coq{LFix}, \coq{GFix} is defined as a setoid, with
an equivalence relation that is the greatest fixpoint of \coq{AppR}.
Additionally, we define smart constructors for the isomorphism of greatest
fixed points
\begin{coqcode}
g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
\end{coqcode}
\dcas{The explanation below is weird ... We have not defined 'ana', but state
their universal property?}
What witnesses the fact that this is the greatest fixed-point is the universal
property of the anamorphism. For a given \coq{F}-coalgebra this is the unique
map that runs the seed function and composes the information from the first step
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In our Coq formalisation, we define anamorphisms as the
unique \coq{F}-coalgebra homomorphisms, that are maps of the following type:
\begin{coqcode}
ana : forall `{F : Cont Sh P} `{setoid A}, Coalg F A ~> A ~> GFix F
\end{coqcode}

>>>>

\subsection{Coalgebras and Anamorphisms}
\label{sec:coalg}
Algebras and catamorphisms dualise nicely to the coinductive setting.  The dual
of an algebra is a coalgebra. A coalgebra can be thought of as an
\emph{observation} map. For example, for a type of states \coq{X} and an
alphabet type \coq{L} we can define a labelled transition system (LTS)
on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{FX = L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.

In general, for a functor \coq{F}, an \coq{F}-coalgebra is a pair of a
type \coq{X} called the \emph{carrier} of the coalgebra and a structure
map \coq{X -> FX}.

In our development we use the following notation for coalgebras:
\begin{coqcode}
Notation Coalg F A := (A ~> App F A).
\end{coqcode}

Dually to the initial \coq{F}-algebra, a final \coq{F}-coalgebra is the greatest
fixed-point for a functor \coq{F}. We define it using a coinductive data type:
\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix-in} is its
inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
defined as a setoid, with an equivalence relation that is the greatest fixpoint
of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
greatest fixed points
\begin{coqcode}
g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
\end{coqcode}

The greatest fixed-point is a terminal \coq{F}-coalgebra in the sense that it
yields a coinductive recursion scheme. Specifically, for any
other \coq{F}-coalgebra there exists a \emph{unique} map, called the
\emph{unfold} or \emph{anamorphism}, such that it applies the observation map,
corecursively generates the rest of the computation and composes the result of
the corecursive call by using the algebra \coq{GFix_in}. In other words, for any
given \coq{F}-coalgebra there exists a unique \coq{F}-coalgebra homomorphism
into the terminal \coq{F}-coalgebra.

% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
% \[\begin{tikzcd}
%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
% 	\arrow["a", from=1-1, to=1-2]
% 	\arrow["b", from=1-2, to=1-3]
% \end{tikzcd}\]
% Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
% carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
% an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.

\mpav{David, I need full def of this ana}
We define anamorphisms as follows:
\begin{coqcode}
ana : forall `{F : Cont Sh P} `{setoid A}, Coalg F A ~> A ~> GFix F
\end{coqcode}
From this definition the universality property falls out:
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In words, for any \coq{F}-coalgebra, if there is any other function \coq{f} that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the same coalgebra.
